{CPU-Scheduling}

### 2024-12-30 LaTeX/Sources/Advanced.tex
finished correction

In this chapter we will look at proportional-share scheduling. This type of policy works with weights and assigns them to the processes. For us these weights will take the form of tickets.
Tickets are a form of currency for the computer. Think of it like money.
These can be handed out by the scheduler and give a process the right to use a resource.
Usually, the rigth to a resource is reflected proportianally by the amount of tickets.
Meaning a higher share of tickets will lead to a higher amount of CPU time.
Logically, the response time is inversely proportional. The more tickets you have the less time you have to wait until you run.
These tickets enable easy comparison between processes.
One important difference between money in real life and tickets is that tickets are not consumed when used.
Practically this means that if you buy something you won't lose your bill.
This property leads to the tickets representing the share of the CPU that the job has.
Higher priority processes are weighted more heavily, meaning that they get more tickets

One could imagen these tickets like reservation places in a restaurant. For now it does not matter how the restaurant assigns the places, like it does not matter how the scheduler policy implements the transition from tickets to CPU time. The only thing that is important is that in the end you will have your places and can eat at the restaurant.

@@

You can do two notable things with tickets. First you can transfer them. 
This does exactly what it means. You decrease your amount of tickets and increase someone else's. 
With this you could temporarily boost other processes.
Say, for example, you have a server and a client.
If the client needs something from the server, then it could temporarily boost the servers resources by transfering some of his own tickets.
Of course this requires that the server is trusted or else it could just scam you out of your tickets.
The second way to operate with tickets is ticket inflation and deflation. This can only be done by the owner of the currency and it is only something that we will look at the next section.

@@

Like other things in computer science the name already says a lot. 
Here randomization is exploited to achieve the proportianal share in runtime.
The concept is, that even if over a small scale there will be unavoidable errors, the longer the scheduler runs the more accurate it becomes.
Think of the tickets as lottery tickets.
If the randomization is fair, than the more tickets you have the more you will win.
There is a downside though.
The randomizer algorithm can get pretty expensive.
There is a way to reuse one random seed though, but we will only quickly touch upon it later.
For an example implementation in C take a look at listing \ref{code:lottery-sched}, which is written by Carl A. Waldspurger \cite{waldspurger95}.

@@

  /* initialize client withspecified allocation */
  void client_init(client_t c, int tickets) {
    c->tickets = tickets; /* initialize client state */ 
    global_tickets += tickets; /* update global sum */
    list_insert(list, c); /* join competition for resource */
  }

@@

Even though it is easy to implement, basing the datastructure on a list is quite inefficient, because of the long search time.
Everytime you want to reschedule, you have to create a random number and loop through the list.
However you could use a balanced binary tree to make it much faster.
This is because a list search time scales with n. 
Therefore if you add one more element you will have to search through one more element.
In contrast a binary search scales not with n.
To be more precise it takes $O(log(n))$ to find a specific value.
On average, with each decision we can eliminate half of the remaining candidates.
Therefore overall it is much faster, especially the more values the list contains.

@@

\subsubsection{Multiwinner Lottery}

Multiwinner Lottery Scheduling is a semi-randomized, semi-deterministic scheduling method.
It is a method to try to minimize the cost of the randomization algorithm.
The basic idea is to reuse the winner.
After you select your winner you take a specific offset to determine the other winners.
One of these lotteries is called a superquantum.

@@

To understand the policy better, let's loot at the example in figure \ref{fig:stride-scheduling}.
We have three processes. Process A has a stride of 160, Process B 320 and Process C has 30. At $t_0$ process A runs. Afterwards its B's turn. C has a much smaller stride, that's why it will have more turns before it catches up to A.
After a while it gets to cyclic pattern.

@@

To solve this, one needs to introduce a constant global pass value, because otherwise the joining processes will monopolize the CPU. 
Look at figure \ref{fig:stride-scheduling} for example.
If a process D joins at the last time interval, than it has to catch up to atleast C. During that time it will have a unreasonable priority over everybody else.
A promised global pass value would eliminate the problem, because one can adjust the entry pass value according to the current situation.
If a task goes into a blocked state or leaves the competition temporarily and than rejoins than its pass value will be global stride plus the stride of the process. The same goes for a brand new job that joins.

>>>>

In this chapter we will look at proportional-share scheduling. This type of policy works with weights and assigns them to the processes. For us these weights will take the form of tickets.
Tickets are a form of currency for the computer. Think of it like money.
These can be handed out by the scheduler and give a process the right to use a resource.
Usually, the right to a resource is reflected proportionally by the amount of tickets.
Meaning a higher share of tickets will lead to a higher amount of CPU time.
Logically, the response time is inversely proportional. The more tickets you have the less time you have to wait until you run.
These tickets enable easy comparison between processes.
One important difference between money in real life and tickets is that tickets are not consumed when used.
Practically this means that if you buy something you won't lose your bill.
This property leads to the tickets representing the share of the CPU that the job has.
Higher priority processes are weighted more heavily, meaning that they get more tickets.

One could imagine these tickets like reservation places in a restaurant. For now it does not matter how the restaurant assigns the places, like it does not matter how the scheduler policy implements the transition from tickets to CPU time. The only thing that is important is that in the end you will have your places and can eat at the restaurant.

@@

You can do two notable things with tickets. First you can transfer them. 
This does exactly what it means. You decrease your amount of tickets and increase someone else's. 
With this you could temporarily boost other processes.
Say, for example, you have a server and a client.
If the client needs something from the server, then it could temporarily boost the servers resources by transferring some of his own tickets.
Of course this requires that the server is trusted or else it could just scam you out of your tickets.
The second way to operate with tickets is ticket inflation and deflation. This can only be done by the owner of the currency and it is only something that we will look at the next section.

@@

Like other things in computer science the name already says a lot. 
Here randomization is exploited to achieve the proportional share in runtime.
The concept is, that even if over a small scale there will be unavoidable errors, the longer the scheduler runs the more accurate it becomes.
Think of the tickets as lottery tickets.
If the randomization is fair, than the more tickets you have the more you will win.
There is a downside though.
The randomizer algorithm can get pretty expensive.
There is a way to reuse one random seed though, but we will only quickly touch upon it later.
For an example implementation in C take a look at listing \ref{code:lottery-sched}, which is written by Carl A. Waldspurger \cite{waldspurger95}.

@@

  /* initialize client with specified allocation */
  void client_init(client_t c, int tickets) {
    c->tickets = tickets; /* initialize client state */ 
    global_tickets += tickets; /* update global sum */
    list_insert(list, c); /* join competition for resource */
  }

@@

Even though it is easy to implement, basing the datas tructure on a list is quite inefficient, because of the long search time.
Every time you want to reschedule, you have to create a random number and loop through the list.
However you could use a balanced binary tree to make it much faster.
This is because a list search time scales with n. 
Therefore if you add one more element you will have to search through one more element.
In contrast a binary search scales not with n.
To be more precise it takes $O(log(n))$ to find a specific value.
On average, with each decision we can eliminate half of the remaining candidates.
Therefore overall it is much faster, especially the more values the list contains.

@@

\subsubsection{Multi-Winner Lottery}

Multi-winner Lottery Scheduling is a semi-randomized, semi-deterministic scheduling method.
It is a method to try to minimize the cost of the randomization algorithm.
The basic idea is to reuse the winner.
After you select your winner you take a specific offset to determine the other winners.
One of these lotteries is called a super-quantum.

@@

To understand the policy better, let's loot at the example in figure \ref{fig:stride-scheduling}.
We have three processes. Process A has a stride of 160, Process B 320 and Process C has 30. At $t_0$ process A runs. Afterwards it's B's turn. C has a much smaller stride, that's why it will have more turns before it catches up to A.
After a while it gets to cyclic pattern.

@@

To solve this, one needs to introduce a constant global pass value, because otherwise the joining processes will monopolize the CPU. 
Look at figure \ref{fig:stride-scheduling} for example.
If a process D joins at the last time interval, then it has to catch up to at least C. During that time it will have a unreasonable priority over everybody else.
A promised global pass value would eliminate the problem, because one can adjust the entry pass value according to the current situation.
If a task goes into a blocked state or leaves the competition temporarily and than rejoins than its pass value will be global stride plus the stride of the process. The same goes for a brand new job that joins.

### 2024-12-30 LaTeX/Sources/Basics.tex
finished correction

The two formulas above are for a single process. To get a feeling of the overall system, we will look at the average. This average highly depends on the system, because the length of the jobs scales with the average turnaround time proportionally.
If on average there are more items for the cashier to scan, then the average time until he is finished will go up.
Therefore we’ll have to look at the measurements for a predetermined set of jobs.
If we do not have the same system, we can not compare the policies.

@@

During the next few chapters and sections we'll assume that one knows the amount of time that a process will take to finish, also known as burst time.
Burst time is the time that a process has to run until it is finished.
Like bursting out of joy, when you finally finish your weekly shopping.
In a real world this is almost impossible\footnote{Here I mean knowning the burst time and not escaping the supermarket.}, unless you can time travel.
One of the most straightforward policy is called first come, first served.
This is what we usually suffer through in the queue to pay.
The policy disregards how many item one has in the cart.
The only important attribute is the time you arrive.
The sooner the better.
In the world of computer science, the policy is better refered to as \emph{First In, First Out (FIFO)}.
Take a look at listing \ref{code:fifo} for a basic python implementation.

@@

\subsection{Shorterst Job First}

As the name already says, the shortest job will be run first.
In order to achieve this, we assume that the cashier knows, who has the least amount of items.
This buyer will be then handeled before anybody else. As you can already tell, this is not really a fair way to treat your customers.
Still if it comes down to pure turnaround time it is in most cases far better\footnote{With the assumption that the tasks arrive at the same time.} than the basic \emph{First In, First Out}.
Waiting for shortest jobs to finish first is far less severe for longer jobs, because they already have a long burst time. However making the short processes wait out the long ones increases the turnaround time by a lot.
For example let's take process A with a burst time of 10 second and process B with 100 seconds. In case of letting A run first we will have $T_A = 10$ and $T_B = 110$.
Therefore, the average turnaround time is:
$$\frac{10 + 110}{2} = \frac{120}{2} = 60 \text{ seconds}$$
In contrast to that if we let process B run first, we will have the turnaround time of 100 and 110. Therefore, the average is:
$$\frac{110 + 100}{2} = \frac{210}{2} = 105 \text{ seconds}$$

@@

As already mentionned in the section \ref{sec:first-lines-of-code}, python is not really meant for writing low level code and therefore it should only be looked at as a simple way of expressing logic.
Even though there were some major improvements, the fact remains that if a client is too late then it will have to wait at least  until the previous shortest customer finishes. This can take a long time and it would be in our best interest to give these newcomers a fair chance. 
It would be great if a rescheduling happens, every time someone joins the queue.

@@

In this section our cashier receives a significant upgrade: it can save the state of one buyer and switch to another one.
In technical terms a policy that can do that is called a preemtive policy. 
The switching itself is called a content switch.
While previously, if another customer came with just a single item, it would not get the privilege of cutting infront of the line. 
He would only receive the priority boost, if the current person is finished.
However, now with the preemptive capabilities the cashier can just save and put aside the people.
This feature is not only useful if there is someone that has a shorter time to completion.
Imagine that someone forgot to put something into their basket.
Now they desparately send back their kid to get that item. While he returns the cashier can save the state and do a context switch.
While processes wait for I/O, others can run in their place.
This waiting state is called being blocked and the technique of running another process is overlapping.
At this stage it does not really make sense to present some code, because it is just too out of place, however, in theory, one would do the sorting of the processes everytime a job finishes, gets blocked or another process joins.

>>>>

The two formulas above are for a single process. To get a feeling of the overall system, we will look at the average. This average highly depends on the system, because the length of the jobs scales with the average turnaround time proportionally.
If on average there are more items for the cashier to scan, then the average time until he is finished will go up.
Therefore we’ll have to look at the measurements for a predetermined set of jobs.
If we do not have the same system, we cannot compare the policies.

@@

During the next few chapters and sections we'll assume that one knows the amount of time that a process will take to finish, also known as burst time.
Burst time is the time that a process has to run until it is finished.
Like bursting out of joy, when you finally finish your weekly shopping.
In a real world this is almost impossible\footnote{Here I mean knowing the burst time and not escaping the supermarket.}, unless you can time travel.
One of the most straightforward policy is called first come, first served.
This is what we usually suffer through in the queue to pay.
The policy disregards how many item one has in the cart.
The only important attribute is the time you arrive.
The sooner the better.
In the world of computer science, the policy is better referred to as \emph{First In, First Out (FIFO)}.
Take a look at listing \ref{code:fifo} for a basic python implementation.

@@

\subsection{Shortest Job First}

As the name already says, the shortest job will be run first.
In order to achieve this, we assume that the cashier knows, who has the least amount of items.
This buyer will be then handled before anybody else. As you can already tell, this is not really a fair way to treat your customers.
Still if it comes down to pure turnaround time it is in most cases far better\footnote{With the assumption that the tasks arrive at the same time.} than the basic \emph{First In, First Out}.
Waiting for shortest jobs to finish first is far less severe for longer jobs, because they already have a long burst time. However making the short processes wait out the long ones increases the turnaround time by a lot.
For example let's take process A with a burst time of 10 second and process B with 100 seconds. In case of letting A run first we will have $T_A = 10$ and $T_B = 110$.
Therefore, the average turnaround time is:
$$\frac{10 + 110}{2} = \frac{120}{2} = 60 \text{ seconds}$$
In contrast to that if we let process B run first, we will have the turnaround time of 100 and 110. Therefore, the average is:
$$\frac{110 + 100}{2} = \frac{210}{2} = 105 \text{ seconds}$$

@@

As already mentioned in the section \ref{sec:first-lines-of-code}, python is not really meant for writing low level code and therefore it should only be looked at as a simple way of expressing logic.
Even though there were some major improvements, the fact remains that if a client is too late then it will have to wait at least  until the previous shortest customer finishes. This can take a long time and it would be in our best interest to give these newcomers a fair chance. 
It would be great if a rescheduling happens, every time someone joins the queue.

@@

In this section our cashier receives a significant upgrade: it can save the state of one buyer and switch to another one.
In technical terms a policy that can do that is called a preemptive policy. 
The switching itself is called a content switch.
While previously, if another customer came with just a single item, it would not get the privilege of cutting in front of the line. 
He would only receive the priority boost, if the current person is finished.
However, now with the preemptive capabilities the cashier can just save and put aside the people.
This feature is not only useful if there is someone that has a shorter time to completion.
Imagine that someone forgot to put something into their basket.
Now they desperately send back their kid to get that item. While he returns the cashier can save the state and do a context switch.
While processes wait for I/O, others can run in their place.
This waiting state is called being blocked and the technique of running another process is overlapping.
At this stage it does not really make sense to present some code, because it is just too out of place, however, in theory, one would do the sorting of the processes every time a job finishes, gets blocked or another process joins.
